Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-03-06 text/front_end_v3.qmd
working through front end. again.

As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. @fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, consisting of those that respond to an initial soliciation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media advertisements designed to solicit inquiries and applications [@RN4895].

@@

A growing college access literature in sociology analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly white public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].

By focusing on the behaviors of individual colleges, scholarship on college access implicitly states that recruiting -- and enrollment management more broadly -- is something done by individual colleges. This statement raises the question, why does recruiting behavior look so similar across colleges of a particular "type." Institutional theory defines the organizational field as "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services and products" [@RN527, p. 143]. The broader enrollment management (EM) industry consists of university personnel (e.g., admissions counselors, VP for enrollment management), professional associations (e.g., National Association for College Admission Counseling), and third-party servicers that produce market research, software, and consulting services (e.g., College Board, ACT, EAB, Slate, PowerSchool). We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Enrollment management is often described as an opaque industry because obtaining concrete information about EM practices is difficult. This is why the title of @RN5024 is *Lifting the Veil on Enrollment Management*. Although sociologists have hinted at the ways enrollment management contributes to inequality in colege access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. More recently, the EM consulting business model is based on software-as-service products (e.g., for example, EAB's [Enroll 360](https://eab.com/solutions/enroll360/) product) [@list_biz]. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. <!-- In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].
-->

The EM industry puts great stock in market research, in part because the EM profession emerged from marketing and market research [@RN4986; @RN2771]. @RN4986 [p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].[^1] In recruiting, market research informs decisions about which prospective students to target with recruiting interventions.<!-- THIS IDEA HAS TO GO SOMEWHERE: IN THE HOOK AT END OF THIS SECTION. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], structuring college access pathways in ways that could amplify or diminish historic inequality in college access.  --> How to acquire market research is another make-or-buy decision. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In 1983, College Board published *Applying Market Research in College Admissions* [@RN4986], which became the basis for their Admitted Student Questionnaire (ASQ) product, and published *The Structure of College Choice* [@RN4982], which became the basis for the Enrollment Planning Service (EPS) product/division. Today, EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Market research often distills large amounts of data into simpler indicators. Therefore, the quantification literature provides a useful lens to understand market research.

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."

@@

@RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402). They introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Speaking to *reactivity* effects, rankings had the intended effect of affecting where students apply and enroll \[CITE\]. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerinags [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK League Tables by @RN5005 show how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school "league tables," which ranked schools based on student performance on high-stakes national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools.<!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

@@

**Correlation**. <!-- In the literature on quantification from critical data studies, --> The discussions of correlation and homophily by @RN4975 introduce ideas salient to the analysis the Market Segment Model and College Board Geomarkets. Correlation measures the extent to which two or more variables move together.<!--The Pearson Correlation Coefficient was developed by the 19th Century statisticians and "biometric eugenicists" Francis Galton and his student Karl Pearson [@RN4975]. --> @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'" Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes.

Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated to a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

Several studies show that using correlational models make predictions reproduces structural inequality [@RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. These correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon whereby has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].[^2]

@@

**Homophily**. @RN4975 argues that homophily is central to models that use correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together." Homophily is a core assumption for network science [@mcpherson2001birds], in which actors -- in this manuscript, high school students -- are linked to one another through direct and indirect network ties -- submitting SAT scores to the same college. Because network science models often draws from rational choice theory, they assume that homophily is the result of voluntary action by individuals, thereby "eras\[ing\] historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks (e.g., Faceboo, X, TikTok), homophily is more than an assumption; homophily is programmed into the algorithms that create connections between users.[^3]

@@

Within the market research profession, homophily is central to geodemography and the creation of market segments. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are targeted to a vertical hierarchy of consumers. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

Negative consequences of homophily
Chun (2021) discusses the negative consequences arising from homophily.Second, when we “engineer homophily” – through algorithms or policies – we make it less likely for heterophily to emerge, eliminating possibilities – of friendships, of matches – that may have occurred in the absence of such engineering.
-->

@@

Reviewing scholarship on quantification, @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. U.S. law school rankings [@RN5003] and UK school league tables [@RN5005] exemplify this sort of salience. Studies of quantification in the sociology of education tend to study consumer-facing metrics. Law school rankings can be conceived as a business-to-consumer product because the goal is to inform consumers (prospective law school students) and consumers pay for the product. By contrast, market research helps businesses identify/target customers and can be conceived as a business-to-business (B2B) product because businesses pay for market research. To our knowledge, scholarship on quantification in education has not examined how third-party market research classifies consumers. Additionally, scholarship on college access has failed to analyze third-party products that colleges use to identify and target prospective students. This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service.

Scholarship on quantification suggests the mechanisms by which market research reproduces structural inequality [CITE CHUN; SOME SOCIOLOGY]. Based on a snapshot of existing social stratification, market research matches verticially categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification. However, this study provides inisight about the mechanism underlying the effect. By analyzing and simulating *Student Search Service* purchases that filter on Geomarkets we show how Geomarkets reproduce historical race-based inequality in college access [THIS LAST BIT FEELS LIKE THE GENERAL CONTRIBUTION TO SOCIOLOGY].

@@

**Creating the Market Segment Model.** In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. The researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in these analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the *Comprehensive Undergraduate Enrollment Planning Project (CUEPP)*. @RN4982 [p. 4] write that,

>>>>

<!-- la la -->
Enrollment management is simultaneously a profession, an administrative structure, and an industry. As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  -->

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, consisting of those that respond to an initial soliciation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media advertisements designed to solicit inquiries and applications [@RN4895].

@@

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. Recruiting visits to high schools are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly white public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

By focusing on the behaviors of individual colleges, scholarship on EM implicitly states that recruiting is something done by individual colleges. 
In addition to university personnel  (e.g., admissions counselors, VP for enrollment management), the EM industry incorporates external stakeholders in the organizational field, including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges. We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways enrollment management contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis.
<!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  -->
<!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].  In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

Although the enrollment management industry is increasingly characterized by software-as-service products sold by private equity backed firms (e.g., for example, EAB's [Enroll 360](https://eab.com/solutions/enroll360/) product) [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. The Student Search Service, created in 1972, became the ubiquitous means of identifying prospects BLAH BLAH. Second, EPS, launcched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification. Student Search Service includes or excludes prospects based on selected attributes while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. THEREFORE WE REVIEW SCHOLARSHIP ON QUANTIFICATION.

<!--

The EM industry puts great stock in market research, in part because the EM profession emerged from marketing and market research [@RN4986; @RN2771]. @RN4986 [p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].[^1] In recruiting, market research informs decisions about which prospective students to target with recruiting interventions. How to acquire market research is another make-or-buy decision. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In 1983, College Board published *Applying Market Research in College Admissions* [@RN4986], which became the basis for their Admitted Student Questionnaire (ASQ) product, and published *The Structure of College Choice* [@RN4982], which became the basis for the Enrollment Planning Service (EPS) product/division. Today, EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Market research often distills large amounts of data into simpler indicators. Therefore, the quantification literature provides a useful lens to understand market research.

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->

@@

@RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402). Reacivity, discipline, and authority are three interrelated themes that describe the effects of quantification. *Reactivity* is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification *disciplines* actors to react in particular ways. Quantification changes decision-making power *authority*, often weakening the discretion of local decision-makers. Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll \[CITE\]. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerinags [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools.<!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

@@

**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4849; @RN4772; ]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and College Board Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"^[Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?']

A bevy of studies show that predictions based on correlations reproduce structural inequality [for a review see @RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. The observed correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon whereby has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].[^2] As we discuss below, @RN4982 analyzed 1980 score-sending behavior (training data) and inferred that demand for college is a function of class. They then recommended that selective colleges recruit from localities that have a high share of affluent, college-educated households.

@@

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science (writ large), in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). As discussed below, @RN4982[chapter 4] *The company We Keep: Colleges and Their Competition* conducts a network analysis based on 1980 SAT score sending behavior whereby two colleges are defined as competitors if a large number of students send SAT scores to both colleges. Because network science models often draws from rational choice theory, @RN4975 argues, it assumes that homophily is the result of voluntary action by individuals.

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Like correlational models discussed above, network science models "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. <!--"Homophily assumes and creates segregation" [@RN4975, p. 95] writes. "It transforms individuals into 'neighbors' who naturally want to live with people 'like them'." --> In commercial social networks, @RN4975 argues, homopphily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar processes in college access because higher education is an industry in which providers are evaluated based on the characteristics of their customers [@RN1549; @RN1237]. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

Homophily is also central to products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are targeted to a vertical hierarchy of consumers. We observe similar processes in our case study, whereby @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there.
<!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

Negative consequences of homophily
Chun (2021) discusses the negative consequences arising from homophily.Second, when we “engineer homophily” – through algorithms or policies – we make it less likely for heterophily to emerge, eliminating possibilities – of friendships, of matches – that may have occurred in the absence of such engineering.

Reviewing scholarship on quantification, @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. U.S. law school rankings [@RN5003] and UK school league tables [@RN5005] exemplify this sort of salience. Studies of quantification in the sociology of education tend to study consumer-facing metrics. Law school rankings can be conceived as a business-to-consumer product because the goal is to inform consumers (prospective law school students) and consumers pay for the product.

-->

@@

**Contribution**. As @RN3519 demonstrates in the chapter *Numbers*, enrollment management is fundamentally concerned with quantification. In the case of college rankings, sociology made visible how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges who are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375]. Beyond rankings, sociological analyses of enrollment management focus on the behavior of colleges and their agents, yielding the implicit assumption that inequalities creates by enrollment management are a function of individual college behavior. We argue that third-party vendors and products structure the enrollment management behavior of colleges. These products take a snapshot of student demand -- without considering structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches verticially categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. These products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociologists.

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification. However, this study provides inisight about the mechanism underlying the effect. By analyzing and simulating *Student Search Service* purchases that filter on Geomarkets we show how Geomarkets reproduce historical race-based inequality in college access [THIS LAST BIT FEELS LIKE THE GENERAL CONTRIBUTION TO SOCIOLOGY].

@@

**Creating Geomarkets and the Market Segment Model.** In 1978, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x].^[Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s.] Zemsky reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential" [@RN4982, p. x].

The geographic concepts inscribed in the Market Segment Model were based on conversations with college admissions officers. @RN4982 \[p. 4\] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of colege-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created three geographic [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets." These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982[pp. 11-12] briefly describes the creation of Geomarket borders:
